---
title: "01 - Differential Accessibility ATAC-seq"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(GenomicRanges)
  library(rtracklayer)
  library(reshape2)
  library(igraph)
  library(ggplot2)
  library(limma)
  library(ComplexHeatmap)
  library(circlize)
})
```


# Goal

Identify differentially accessible OCRs and generate ATAC clusters used later in GRN filtering.

# Load ATAC data

```{r load-atac}
atac <- readRDS(file.path("data", "atac_norm_data.rds"))
atac_metadata <- readRDS(file.path("data", "atac_metadata.rds"))
atac_anno <- readRDS(file.path("data", "atac_OCR_annotation.rds"))

dim(atac)
head(atac_metadata)
head(atac_anno)

```


# Differential accessibility analysis from normalized ATAC data

This workflow starts from the normalized ATAC matrix and tests chromatin accessibility changes across B-cell differentiation transitions.

```{r differential-accessibility-normalized-atac}

# Match metadata to ATAC columns
atac_metadata <- atac_metadata[match(colnames(atac), rownames(atac_metadata)), ]

stopifnot(all(rownames(atac_metadata) == colnames(atac)))

# Define cell-type order
atac_metadata$CellType <- factor(
  atac_metadata$CellType,
  levels = c(
    "HSC",
    "CLP",
    "proB",
    "preB",
    "ImmatureB",
    "Transitional_B",
    "Naive_CD5neg",
    "Naive_CD5pos"
  )
)

# Build design matrix
# If donor information is available, include it as a covariate.
if ("Donor" %in% colnames(atac_metadata)) {
  design <- model.matrix(~ 0 + CellType + Donor, data = atac_metadata)
} else if ("Factor" %in% colnames(atac_metadata)) {
  design <- model.matrix(~ 0 + CellType + Factor, data = atac_metadata)
} else {
  design <- model.matrix(~ 0 + CellType, data = atac_metadata)
}

colnames(design) <- gsub("CellType", "", colnames(design))

design
```

# Fit limma model

```{r fit-limma-atac}
# log transform the matrix
fit <- lmFit(atac, design)
```

# Define B-cell differentiation transition contrasts

```{r atac-transition-contrasts}
contrast_matrix <- makeContrasts(
  HSC_CLP = CLP - HSC,
  CLP_ProB = proB - CLP,
  ProB_PreB = preB - proB,
  PreB_ImmatureB = ImmatureB - preB,
  ImmatureB_TransitionalB = Transitional_B - ImmatureB,
  TransitionalB_NaiveCD5neg = Naive_CD5neg - Transitional_B,
  TransitionalB_NaiveCD5pos = Naive_CD5pos - Transitional_B,
  NaiveCD5neg_NaiveCD5pos = Naive_CD5pos - Naive_CD5neg,
  levels = design
)

contrast_matrix
```

# Extract DARs per transition

```{r extract-dars}
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

dar_list <- lapply(colnames(contrast_matrix), function(contrast_name) {
  res <- topTable(
    fit2,
    coef = contrast_name,
    number = Inf,
    sort.by = "none",
    adjust.method = "BH"
  )

  res$peak <- rownames(res)
  res$contrast <- contrast_name

  res
})

names(dar_list) <- colnames(contrast_matrix)

saveRDS(
  dar_list,
  file.path("results", "atac_differential_accessibility_all_contrasts.rds")
)

head(dar_list[[1]])
```

# Build transition matrices

This reproduces the structure of the original transition-level output:

- `sig_*`: direction of significant accessibility change  
  - `1` = increased accessibility
  - `-1` = decreased accessibility
  - `0` = not significant
- `p.adj_*`: adjusted p-value
- `logFC_*`: log fold-change

```{r build-transition-matrices}
all_padj <- do.call(
  cbind,
  lapply(dar_list, function(x) x$adj.P.Val)
)

all_logFC <- do.call(
  cbind,
  lapply(dar_list, function(x) x$logFC)
)

rownames(all_padj) <- rownames(atac)
rownames(all_logFC) <- rownames(atac)

colnames(all_padj) <- paste0("p.adj_", names(dar_list))
colnames(all_logFC) <- paste0("logFC_", names(dar_list))

sig <- matrix(
  0,
  nrow = nrow(atac),
  ncol = length(dar_list),
  dimnames = list(
    rownames(atac),
    paste0("sig_", names(dar_list))
  )
)

sig[all_padj < 0.01 & all_logFC > 0] <- 1
sig[all_padj < 0.01 & all_logFC < 0] <- -1

dea_results <- data.frame(sig, all_padj, all_logFC)

head(dea_results)
```

# Save differential accessibility results

```{r save-dar-results}
write.table(
  dea_results,
  file.path("results", "dea_transitions.txt"),
  sep = "\t",
  quote = FALSE
)

transition_cols <- grep("^sig_", colnames(sig), value = TRUE)

dea_sig <- sig[
  apply(sig[, transition_cols, drop = FALSE], 1, function(x) sum(x != 0)) > 0,
  ,
  drop = FALSE
]

write.table(
  dea_sig,
  file.path("results", "sig_dea_transitions.txt"),
  sep = "\t",
  quote = FALSE
)

dim(dea_sig)
```

# Create downstream ATAC cluster object

The downstream GRN workflow uses `cluster_atac$peak` to retain enhancer OCRs that are differentially accessible in at least one main transition.

```{r create-cluster-atac}
cluster_atac <- data.frame(
  peak = rownames(dea_sig)
)

saveRDS(
  cluster_atac,
  file.path("results", "atac_dar_clustering.rds")
)

head(cluster_atac)
```

# Quick visualization

```{r plot-dar-summary}
dar_summary <- data.frame(
  contrast = names(dar_list),
  n_up = colSums(sig == 1),
  n_down = colSums(sig == -1)
)

dar_summary_long <- dar_summary |>
  pivot_longer(
    cols = c(n_up, n_down),
    names_to = "direction",
    values_to = "n_peaks"
  )

ggplot(dar_summary_long, aes(x = contrast, y = n_peaks, fill = direction)) +
  geom_col(position = "dodge") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Transition",
    y = "Number of DARs",
    fill = "Direction"
  )
```
