---
title: "02 - Differential Expression RNA-seq"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(limma)
  library(ggplot2)
})
```


# Goal

Identify differentially expressed genes and generate RNA clusters used later in GRN filtering.

# Load RNA data

```{r load-rna}
rna <- readRDS(file.path("data", "rna_norm_data.rds"))
rna_metadata <- readRDS(file.path("data", "rna_metadata.rds"))
rna_anno <- readRDS(file.path("data", "rna_gene_annotation.rds"))

dim(rna)
head(rna_metadata)
head(rna_anno)
```

# Goal

Identify differentially expressed genes across B-cell differentiation transitions using normalized RNA-seq data.`

# Match metadata to RNA matrix

```{r match-metadata}
# If metadata has sample IDs as rownames, match directly.
# Otherwise, try to use common sample ID columns.

if (!all(colnames(rna) %in% rownames(rna_metadata))) {

  possible_id_cols <- c("SampleID", "Alias", "sample", "sample_id", "Sample", "ID")

  id_col <- possible_id_cols[
    possible_id_cols %in% colnames(rna_metadata)
  ][1]

  if (is.na(id_col)) {
    stop(
      "Could not match RNA sample names to metadata. ",
      "Please set rownames(rna_metadata) to sample IDs or provide a SampleID/Alias column."
    )
  }

  rownames(rna_metadata) <- rna_metadata[[id_col]]
}

rna_metadata <- rna_metadata[
  match(colnames(rna), rownames(rna_metadata)),
  ,
  drop = FALSE
]

stopifnot(all(rownames(rna_metadata) == colnames(rna)))

head(rna_metadata)
```

# Define cell type and donor factors

```{r define-factors}

rna_metadata$CellType <- factor(
  rna_metadata$CellType,
  levels = c(
    "HSC",
    "CLP",
    "proB",
    "preB",
    "ImmatureB",
    "Transitional_B",
    "Naive_CD5neg",
    "Naive_CD5pos"
  )
)

table(rna_metadata$CellType)
```

# Build design matrix

```{r design-matrix}
# Include donor as covariate if available.
if ("Donor" %in% colnames(rna_metadata)) {

  rna_metadata$Donor <- factor(rna_metadata$Donor)

  design <- model.matrix(
    ~ 0 + CellType + Donor,
    data = rna_metadata
  )

} else if ("Factor" %in% colnames(rna_metadata)) {

  rna_metadata$Factor <- factor(rna_metadata$Factor)

  design <- model.matrix(
    ~ 0 + CellType + Factor,
    data = rna_metadata
  )

} else {

  design <- model.matrix(
    ~ 0 + CellType,
    data = rna_metadata
  )
}

colnames(design) <- gsub("CellType", "", colnames(design))

design
```

# Fit limma model

```{r fit-limma}
fit <- lmFit(rna, design)
fit2 <- eBayes(fit)

whole_stats <- topTable(
  fit2,
  n = nrow(rna),
  sort.by = "none"
)

sum(whole_stats$adj.P.Val < 0.01)
sum(whole_stats$adj.P.Val < 0.05)

head(whole_stats)
```

# Define developmental contrasts

```{r contrasts}
contr.matrix <- makeContrasts(
  HSC_CLP = HSC - CLP,
  CLP_ProB = CLP - proB,
  ProB_PreB = proB - preB,
  PreB_ImmatureB = preB - ImmatureB,
  ImmatureB_TransitionalB = ImmatureB - Transitional_B,
  TransitionalB_NaiveCD5neg = Transitional_B - Naive_CD5neg,
  TransitionalB_NaiveCD5pos = Transitional_B - Naive_CD5pos,
  NaiveCD5neg_NaiveCD5pos = Naive_CD5neg - Naive_CD5pos,
  levels = colnames(design)
)

contr.matrix
```

# Differential expression across transitions

```{r fit-contrasts}
fit.contrasts <- contrasts.fit(
  fit,
  contr.matrix
)

fit.contrasts2 <- eBayes(
  fit.contrasts
)

contrast_stats <- topTable(
  fit.contrasts2,
  number = nrow(rna),
  sort.by = "none"
)

sum(contrast_stats$adj.P.Val < 0.01)
sum(contrast_stats$adj.P.Val < 0.05)

head(contrast_stats)
```

# Build transition matrices

This reproduces the binary transition-level output of the original RNA differential expression script.

- `1` = upregulated
- `-1` = downregulated
- `0` = not significant

Thresholds:

```text
adjusted p-value < 0.05
absolute logFC > 0.58
```

The logFC threshold corresponds approximately to fold-change > 1.5.

```{r transition-matrices}
all_padj <- matrix(
  ncol = ncol(contr.matrix),
  nrow = nrow(rna)
)

all_fch <- matrix(
  ncol = ncol(contr.matrix),
  nrow = nrow(rna)
)

sig <- matrix(
  0,
  ncol = ncol(contr.matrix),
  nrow = nrow(rna)
)

for (i in seq_len(ncol(contr.matrix))) {

  target_stats <- topTable(
    fit.contrasts2,
    coef = i,
    number = nrow(rna),
    sort.by = "none"
  )

  all_padj[, i] <- target_stats$adj.P.Val
  all_fch[, i] <- target_stats$logFC

  sel_pos <- intersect(
    which(all_padj[, i] < 0.05),
    which(all_fch[, i] > 0.58)
  )

  sig[sel_pos, i] <- 1

  sel_neg <- intersect(
    which(all_padj[, i] < 0.05),
    which(all_fch[, i] < -0.58)
  )

  sig[sel_neg, i] <- -1
}

colnames(all_padj) <- paste0(
  "p.adj_",
  colnames(contr.matrix)
)

colnames(all_fch) <- paste0(
  "logFC_",
  colnames(contr.matrix)
)

colnames(sig) <- paste0(
  "sig_",
  colnames(contr.matrix)
)

rownames(all_padj) <- rownames(rna)
rownames(all_fch) <- rownames(rna)
rownames(sig) <- rownames(rna)

dea_results <- data.frame(
  sig,
  all_padj,
  all_fch
)

head(dea_results)
```

# Save full DE results

```{r save-de-results}
write.table(
  dea_results,
  file.path(
    results_dir,
    "dea_results.txt"
  ),
  sep = "\t",
  dec = ".",
  quote = FALSE
)

saveRDS(
  dea_results,
  file.path(
    results_dir,
    "rna_differential_expression_all_transitions.rds"
  )
)
```

# Significant DE genes

```{r significant-de-genes}
is_de <- apply(
  sig,
  1,
  function(x) sum(x != 0)
) > 0

dea_sig <- sig[
  is_de,
  ,
  drop = FALSE
]

dim(dea_sig)

write.table(
  dea_sig,
  file.path(
    results_dir,
    "dea_significant_binary.txt"
  ),
  sep = "\t",
  dec = ".",
  quote = FALSE
)
```

# Create cluster_rna object for GRN analysis

The downstream GRN workflow uses:

```text
cluster_rna$gene
```

to keep TFs and target genes that are differentially expressed across B-cell differentiation.

```{r create-cluster-rna}
cluster_rna <- as.data.frame(
  dea_sig
)

cluster_rna$gene <- rownames(cluster_rna)

saveRDS(
  cluster_rna,
  file.path(
    results_dir,
    "rna_deg_clustering.rds"
  )
)

head(cluster_rna)
```

# Quick visualization

```{r plot-de-summary}
de_summary <- data.frame(
  contrast = colnames(sig),
  n_up = colSums(sig == 1),
  n_down = colSums(sig == -1)
)

de_summary_long <- de_summary |>
  pivot_longer(
    cols = c(n_up, n_down),
    names_to = "direction",
    values_to = "n_genes"
  )

ggplot(
  de_summary_long,
  aes(
    x = contrast,
    y = n_genes,
    fill = direction
  )
) +
  geom_col(position = "dodge") +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  ) +
  labs(
    x = "Transition",
    y = "Number of DE genes",
    fill = "Direction"
  )
```

# Output files

This chapter creates:

```text
results/dea_results.txt
results/dea_significant_binary.txt
results/rna_differential_expression_all_transitions.rds
results/rna_deg_clustering.rds
```

