---
title: "07 - TF Assignment with JASPAR, chromVAR and Precomputed Footprints"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(rtracklayer)
  library(SummarizedExperiment)
  library(chromVAR)
  library(motifmatchr)
  library(TFBSTools)
  library(JASPAR2022)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(Matrix)
  library(ComplexHeatmap)
  library(circlize)
  library(ggplot2)
})
```

# Goal

Assign transcription factors to candidate OCRs using:

1. JASPAR motif evidence.
2. Precomputed footprint evidence.
3. chromVAR TF motif accessibility deviations.

This chapter creates the downstream GRN input:

```text
results/ocr_tf_matrix.txt
```

# Workflow

```text
Candidate CREs / OCRs
        ↓
OCR genomic ranges
        ↓
JASPAR motif matching
        ↓
Precomputed footprint overlap
        ↓
Footprint-supported OCR–TF matrix
        ↓
chromVAR motif deviation scores
        ↓
TF accessibility profiles
        ↓
GRN input
```

# Important note about HINT and Wellington

Footprinting is **not run during the workshop**.

The original workflow used HINT-ATAC and Wellington on an HPC cluster. The outputs were unified and summarized into consensus footprint BED files per cell type.

During the workshop we only load the precomputed BED files. For a full pipeline check https://github.com/TranslationalBioinformaticsUnit/bcell/tree/main

```{r hpc-footprinting-note, eval=FALSE}
# HPC-only preprocessing example.
# Do not run during the workshop.

# HINT-ATAC:
# rgt-hint footprinting \
#   --atac-seq \
#   --paired-end \
#   --organism=hg38 \
#   --output-prefix=sample_footprint \
#   --output-location=footprints/hint \
#   sample.bam \
#   sample_peaks.bed

# Wellington:
# wellington_footprints.py sample_peaks.bed sample.bam footprints/wellington
```

# Define paths

```{r paths}
data_dir <- if (dir.exists("data")) "data" else "."
results_dir <- "results"

cat("Using data directory:", normalizePath(data_dir), "\n")
cat("Using results directory:", normalizePath(results_dir), "\n")
```

# Load candidate CREs

```{r load-candidate-cres}
candidate_CREs <- readRDS(file.path(results_dir, "candidate_CREs.rds"))

head(candidate_CREs)
dim(candidate_CREs)
```

# Build OCR GRanges

Candidate CREs contain OCR IDs in the format `chr_start_end`.

```{r build-ocr-granges}
peak_ids <- unique(candidate_CREs$peak)
peak_split <- strsplit(peak_ids, "_")

ocr_df <- data.frame(
  chr = vapply(peak_split, `[`, character(1), 1),
  start = as.numeric(vapply(peak_split, `[`, character(1), 2)),
  end = as.numeric(vapply(peak_split, `[`, character(1), 3)),
  peak = peak_ids
)

ocr_df <- ocr_df[!is.na(ocr_df$start) & !is.na(ocr_df$end), ]

ocr_gr <- makeGRangesFromDataFrame(
  ocr_df,
  keep.extra.columns = TRUE
)

names(ocr_gr) <- ocr_df$peak
ocr_gr
```

# Load precomputed footprint BED files

```{r load-footprints}
footprint_files <- list.files(
  data_dir,
  pattern = "consensus_unified_footprint.*\\.bed$|footprint.*\\.bed$",
  full.names = TRUE
)

footprint_files <- footprint_files[
  !grepl("all_cell", basename(footprint_files), ignore.case = TRUE)
]

footprint_files
```

```{r import-footprints}
footprints <- lapply(footprint_files, import)

names(footprints) <- basename(footprint_files)
names(footprints) <- gsub("_consensus_unified_footprint.*", "", names(footprints))
names(footprints) <- gsub("\\.bed$", "", names(footprints))
names(footprints) <- gsub("\\(.*\\)", "", names(footprints))

names(footprints)
lengths(footprints)
```

# Create unified footprint set

```{r unified-footprints}
footprints2 <- Map(function(gr, nm) {
  gr$celltype <- nm
  gr
}, footprints, names(footprints))

all_footprints <- reduce(do.call(c, unname(footprints2)))

all_footprints <- all_footprints[width(all_footprints) >= 6]
all_footprints
```

# Identify OCRs overlapping footprints

```{r ocr-footprint-overlap}
ocr_fp_hits <- findOverlaps(
  ocr_gr,
  all_footprints,
  type = "any",
  minoverlap = 1
)

ocr_has_footprint <- names(ocr_gr) %in% names(ocr_gr)[unique(queryHits(ocr_fp_hits))]

table(ocr_has_footprint)
```

# JASPAR motif collection

```{r jaspar-motifs}
opts <- list(
  species = 9606,
  collection = "CORE",
  all_versions = FALSE
)

pfm <- getMatrixSet(
  JASPAR2022,
  opts
)

length(pfm)
pfm[1:3]
```

# Prepare genome sequence information

```{r prepare-genome}
genome <- BSgenome.Hsapiens.UCSC.hg38
standard_chr <- paste0("chr", c(1:22, "X", "Y"))

ocr_gr <- keepSeqlevels(
  ocr_gr,
  intersect(seqlevels(ocr_gr), standard_chr),
  pruning.mode = "coarse"
)

seqlevelsStyle(ocr_gr) <- "UCSC"
ocr_gr
```

# Match JASPAR motifs to OCRs

```{r match-jaspar-motifs}
motif_ix <- matchMotifs(
  pwms = pfm,
  subject = ocr_gr,
  genome = genome,
  out = "matches"
)

motif_matrix <- motifMatches(motif_ix)

dim(motif_matrix)
motif_matrix[1:5, 1:5]
```

# Convert motifs to TF names

```{r motif-to-tf-names}
motif_names <- name(pfm)

tf_names <- motif_names
tf_names <- gsub("::", "_", tf_names)
tf_names <- gsub("\\(var\\.2\\)", "", tf_names)
tf_names <- gsub("\\(var\\.3\\)", "", tf_names)
tf_names <- gsub("\\s+", "", tf_names)

colnames(motif_matrix) <- tf_names

motif_matrix_binary <- as.matrix(motif_matrix)
storage.mode(motif_matrix_binary) <- "numeric"
rownames(motif_matrix_binary) <- names(ocr_gr)

dim(motif_matrix_binary)
```

# Keep only footprint-supported motif calls

The motif matrix tells us where a TF could bind based on sequence. The footprint overlap tells us which OCRs show local evidence of TF occupancy.

```{r footprint-supported-motif-matrix}
ocr_has_footprint <- ocr_has_footprint[match(rownames(motif_matrix_binary), names(ocr_gr))]

ocr_tf_matrix <- motif_matrix_binary
ocr_tf_matrix[!ocr_has_footprint, ] <- 0

ocr_tf_matrix <- ocr_tf_matrix[, colSums(ocr_tf_matrix) > 0, drop = FALSE]
ocr_tf_matrix <- ocr_tf_matrix[rowSums(ocr_tf_matrix) > 0, , drop = FALSE]

dim(ocr_tf_matrix)
```

# Collapse duplicated TF columns

```{r collapse-duplicated-tfs}
if (any(duplicated(colnames(ocr_tf_matrix)))) {

  tf_unique <- unique(colnames(ocr_tf_matrix))

  ocr_tf_collapsed <- matrix(
    0,
    nrow = nrow(ocr_tf_matrix),
    ncol = length(tf_unique),
    dimnames = list(
      rownames(ocr_tf_matrix),
      tf_unique
    )
  )

  for (tf in tf_unique) {
    idx <- which(colnames(ocr_tf_matrix) == tf)
    ocr_tf_collapsed[, tf] <- as.numeric(rowSums(ocr_tf_matrix[, idx, drop = FALSE]) > 0)
  }

  ocr_tf_matrix <- ocr_tf_collapsed
}

dim(ocr_tf_matrix)
```

# Save OCR–TF matrix

```{r save-ocr-tf-matrix}
write.table(
  ocr_tf_matrix,
  file.path(results_dir, "ocr_tf_matrix.txt"),
  sep = "\t",
  quote = FALSE,
  row.names = TRUE,
  col.names = NA
)

saveRDS(
  ocr_tf_matrix,
  file.path(results_dir, "ocr_tf_matrix.rds")
)

ocr_tf_matrix[1:5, 1:min(5, ncol(ocr_tf_matrix))]
```

# Summary of TF assignments

```{r tf-assignment-summary}
tf_summary <- data.frame(
  TF = colnames(ocr_tf_matrix),
  n_OCRs = colSums(ocr_tf_matrix)
) |>
  arrange(desc(n_OCRs))

head(tf_summary, 20)

write.table(
  tf_summary,
  file.path(results_dir, "ocr_tf_assignment_summary.txt"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

```{r plot-tf-summary}
png(
  filename = file.path(results_dir, "tf_assignment_summary.png"),
  width = 1200,
  height = 1200, res = 300
)
tf_summary |>
  slice_head(n = 30) |>
  ggplot(aes(x = reorder(TF, n_OCRs), y = n_OCRs)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    x = "TF",
    y = "Number of footprint-supported OCRs",
    title = "Top TFs by number of assigned OCRs"
  )
dev.off()
```

# Prepare chromVAR input

chromVAR requires **raw or integer-like ATAC counts**, not log-normalized accessibility values.

```{r load-atac-counts}
counts_rds <- file.path(data_dir, "atac_raw_data.rds")

if (file.exists(counts_rds)) {

  atac_counts <- readRDS(counts_rds)

} else if (file.exists(counts_txt)) {

  atac_counts <- as.matrix(
    read.delim(
      counts_txt,
      check.names = FALSE,
      row.names = 1
    )
  )

} else {

  atac_counts <- NULL
  message("No raw ATAC count matrix found. chromVAR section will be skipped.")
}

if (!is.null(atac_counts)) {
  dim(atac_counts)
}
```

# Load ATAC metadata for chromVAR

```{r load-atac-metadata}
metadata_rds <- file.path(data_dir, "atac_metadata.rds")

if (file.exists(metadata_rds)) {

  atac_metadata <- readRDS(metadata_rds)

} else if (file.exists(metadata_txt)) {

  atac_metadata <- read.table(
    metadata_txt,
    sep = "\t",
    header = TRUE,
    check.names = FALSE
  )

} else {

  atac_metadata <- NULL
  message("No ATAC metadata found. chromVAR section will be skipped.")
}

if (!is.null(atac_metadata)) {
  head(atac_metadata)
}
```

# Match counts, peaks and metadata

```{r chromvar-matching}
run_chromvar <- !is.null(atac_counts) && !is.null(atac_metadata)

if (run_chromvar) {

  if (!all(colnames(atac_counts) %in% rownames(atac_metadata))) {

    possible_id_cols <- c("SampleID", "Alias", "sample", "sample_id", "Sample", "ID")

    id_col <- possible_id_cols[
      possible_id_cols %in% colnames(atac_metadata)
    ][1]

    if (!is.na(id_col)) {
      rownames(atac_metadata) <- atac_metadata[[id_col]]
    }
  }

  atac_metadata <- atac_metadata[
    match(colnames(atac_counts), rownames(atac_metadata)),
    ,
    drop = FALSE
  ]

  run_chromvar <- all(rownames(atac_metadata) == colnames(atac_counts))

  if (!run_chromvar) {
    message("Could not match ATAC metadata to count matrix. chromVAR will be skipped.")
  }
}
```

```{r chromvar-peak-matching}
if (run_chromvar) {

  common_peaks <- intersect(rownames(atac_counts), names(ocr_gr))
  common_peaks <- intersect(common_peaks, rownames(ocr_tf_matrix))

  atac_counts_cv <- atac_counts[common_peaks, , drop = FALSE]
  ocr_gr_cv <- ocr_gr[common_peaks]
  ocr_tf_matrix_cv <- ocr_tf_matrix[common_peaks, , drop = FALSE]

  dim(atac_counts_cv)
  dim(ocr_tf_matrix_cv)
}
```

# Run chromVAR

```{r run-chromvar}

fragment_counts <- SummarizedExperiment(
    assays = list(counts = atac_counts_cv),
    rowRanges = ocr_gr_cv,
    colData = atac_metadata
  )

  fragment_counts <- addGCBias(
    fragment_counts,
    genome = genome
  )

keep <- base::rowSums(counts(fragment_counts)) > 0

fragment_counts_filtered <- fragment_counts[keep, ]

  ocr_tf_matrix_cv <- ocr_tf_matrix_cv[
    rownames(fragment_counts_filtered),
    ,
    drop = FALSE
  ]

  annotations <- getAnnotations(
    ocr_tf_matrix_cv,
    rowRanges = rowRanges(fragment_counts_filtered)
  )

  dev <- computeDeviations(
    object = fragment_counts_filtered,
    annotations = annotations
  )

  TF_deviations <- deviations(dev)
  TF_scores <- deviationScores(dev)
  TF_variability <- computeVariability(dev)

  saveRDS(
    dev,
    file.path(results_dir, "chromVAR_TF_deviations_object.rds")
  )

  write.table(
    TF_deviations,
    file.path(results_dir, "TF_TFBS_deviations.txt"),
    sep = "\t",
    quote = FALSE
  )

  write.table(
    TF_scores,
    file.path(results_dir, "TF_TFBS_score.txt"),
    sep = "\t",
    quote = FALSE
  )

  write.table(
    TF_variability,
    file.path(results_dir, "TF_TFBS_variability.txt"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )

  head(TF_variability)

```

# chromVAR sample correlation

```{r chromvar-sample-correlation}
sample_cor <- getSampleCorrelation(dev)

  saveRDS(
    sample_cor,
    file.path(results_dir, "chromVAR_sample_correlation.rds")
  )

Heatmap(
    sample_cor,
    show_row_names = FALSE,
    show_column_names = FALSE,
    name = "rho",
    column_title = "chromVAR sample correlation"
  )

```

# chromVAR TF score heatmap

```{r chromvar-tf-heatmap}
top_variable_tfs <- TF_variability |>
    arrange(desc(variability)) |>
    slice_head(n = min(50, nrow(TF_variability))) |>
    pull(name)

  TF_scores_top <- TF_scores[
    top_variable_tfs,
    ,
    drop = FALSE
  ]

Heatmap(
    TF_scores_top,
    show_row_names = TRUE,
    show_column_names = FALSE,
    name = "Z-score",
    column_title = "Top variable TF motif accessibility scores"
  )
```

# Optional: correlate chromVAR TF scores with RNA expression

```{r tfbs-rna-correlation}
rna_file <- file.path(data_dir, "rna_norm_data.rds")
rna_anno_file <- file.path(data_dir, "rna_gene_annotation.rds")

if (
  run_chromvar &&
  file.exists(rna_file) &&
  file.exists(rna_anno_file)
) {

  rna <- readRDS(rna_file)
  rna_anno <- readRDS(rna_anno_file)

  ids_atac_rna <- intersect(
    colnames(TF_scores),
    colnames(rna)
  )

  if (length(ids_atac_rna) > 2) {

    rna2 <- as.matrix(rna[, ids_atac_rna, drop = FALSE])
    TF_scores2 <- as.matrix(TF_scores[, ids_atac_rna, drop = FALSE])

    if ("hgnc_symbol" %in% colnames(rna_anno)) {
      rna_symbols <- rna_anno$hgnc_symbol[
        match(rownames(rna2), rna_anno$ensembl_gene_id)
      ]

      keep <- !is.na(rna_symbols) & rna_symbols != ""
      rna2 <- rna2[keep, , drop = FALSE]
      rownames(rna2) <- rna_symbols[keep]
    }

    common_tfs <- intersect(
      rownames(TF_scores2),
      rownames(rna2)
    )

    tfbs_rna_correlation <- lapply(common_tfs, function(tf) {

      cc <- cor.test(
        as.numeric(TF_scores2[tf, ]),
        as.numeric(rna2[tf, ]),
        method = "spearman"
      )

      data.frame(
        TF = tf,
        spearman_pval = cc$p.value,
        spearman_rho = as.numeric(cc$estimate)
      )
    }) |>
      bind_rows()

    tfbs_rna_correlation$spearman_adj_pval <- p.adjust(
      tfbs_rna_correlation$spearman_pval,
      method = "bonferroni"
    )

    write.table(
      tfbs_rna_correlation,
      file.path(results_dir, "TFBS_RNA_correlation.txt"),
      sep = "\t",
      quote = FALSE,
      row.names = FALSE
    )

    tfbs_rna_correlation |>
      arrange(spearman_adj_pval) |>
      head(20)
  }
}
```

# Output files

This chapter creates:

```text
results/ocr_tf_matrix.txt
results/ocr_tf_matrix.rds
results/ocr_tf_assignment_summary.txt
results/chromVAR_TF_deviations_object.rds
results/TF_TFBS_deviations.txt
results/TF_TFBS_score.txt
results/TF_TFBS_variability.txt
results/chromVAR_sample_correlation.rds
results/TFBS_RNA_correlation.txt
```

# Notes for downstream GRN reconstruction

The next chapter expects:

```text
results/ocr_tf_matrix.txt
```

This matrix has:

- rows = OCRs
- columns = TFs
- values = binary TF assignment

Only OCRs with both motif evidence and footprint support are retained.
