---
title: "10 - Network Analysis with igraph"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(rtracklayer)
  library(reshape2)
  library(igraph)
  library(ggplot2)
})
```


# Goal

Build GRNs and compute network properties:
- degree [total number of edges (connections) that a TF has associated] 
- betweenness [influence of a given TF over the flow of information in the network] centrality measures computed within each cell-type specific GRN
- edges [connections between TFs and genes, weighted by the strength of their correlation (rho)]
- nodes [TFs and genes in the network].

# Load inputs

```{r load-inputs}
tf_ocr_gene_extended <- readRDS(
  file.path("results", "tf_ocr_gene_interactions_curated_extended.rds")
)

cluster_rna <- readRDS(file.path("results", "rna_deg_clustering.rds"))
```

# Helper function

```{r centrality-function}
add_centrality <- function(net) {
  V(net)$degree <- igraph::degree(net)
  V(net)$indegree <- igraph::degree(net, mode = "in")
  V(net)$outdegree <- igraph::degree(net, mode = "out")
  V(net)$eig <- eigen_centrality(net, directed = TRUE)$vector
  V(net)$hubs <- hub_score(net)$vector
  V(net)$authorities <- authority_score(net)$vector
  V(net)$closeness <- closeness(net)
  V(net)$betweenness <- betweenness(net)

  net
}
```

# Global B-cell GRN

```{r global-grn}
tf_gene <- unique(data.frame(
  tf_ensembl = tf_ocr_gene_extended$tf_ensembl,
  gene_ensembl = tf_ocr_gene_extended$gene_ensembl,
  gene_tf_rho = tf_ocr_gene_extended$gene_tf_rho
))

edges <- data.frame(
  from = tf_gene$tf_ensembl,
  to = tf_gene$gene_ensembl,
  weight = abs(tf_gene$gene_tf_rho)
)

nodes <- data.frame(
  id = unique(c(edges$from, edges$to)),
  label = unique(c(edges$from, edges$to))
)

net <- graph_from_data_frame(edges, nodes, directed = TRUE)
net <- add_centrality(net)

net
```

# Node properties

```{r global-node-properties}
global_node_properties <- data.frame(
  gene = V(net)$name,
  degree = V(net)$degree,
  indegree = V(net)$indegree,
  outdegree = V(net)$outdegree,
  eigenvector = V(net)$eig,
  hub_score = V(net)$hubs,
  authority_score = V(net)$authorities,
  closeness = V(net)$closeness,
  betweenness = V(net)$betweenness
)

global_node_properties |>
  arrange(desc(outdegree)) |>
  head(20)
```

# Cell-type-specific GRNs

```{r cell-type-networks}
cell_type <- c(
  "HSC",
  "CLP",
  "proB",
  "preB",
  "ImmatureB",
  "Transitional_B",
  "Naive_CD5neg",
  "Naive_CD5pos"
)

list_grn <- vector("list", length(cell_type))
names(list_grn) <- cell_type

for (ct in cell_type) {
  interaction_col <- paste0(ct, "_interaction")

  aa <- tf_ocr_gene_extended[
    tf_ocr_gene_extended[[interaction_col]] == "yes",
  ]

  if (ct %in% colnames(cluster_rna)) {
    exp_genes <- rownames(cluster_rna[cluster_rna[[ct]] == 1, , drop = FALSE])

    aa <- aa[
      aa$tf_ensembl %in% exp_genes &
        aa$gene_ensembl %in% exp_genes,
    ]
  }

  edges <- unique(data.frame(
    from = aa$tf_ensembl,
    to = aa$gene_ensembl,
    weight = abs(aa$gene_tf_rho)
  ))

  if (nrow(edges) == 0) {
    list_grn[[ct]] <- make_empty_graph(directed = TRUE)
    next
  }

  nodes <- data.frame(
    id = unique(c(edges$from, edges$to)),
    label = unique(c(edges$from, edges$to))
  )

  ct_net <- graph.data.frame(edges, nodes, directed = TRUE)
  ct_net <- add_centrality(ct_net)

  list_grn[[ct]] <- ct_net
}

list_grn
```

# Compare networks

```{r network-summary}
network_summary <- data.frame(
  cell_type = names(list_grn),
  nodes = sapply(list_grn, vcount),
  edges = sapply(list_grn, ecount)
)

network_summary
```

# Save networks

```{r save-networks}
saveRDS(list_grn, file.path("results", "bcell_grn_by_cell_type.rds"))
save(net, file = file.path("results", "bcell_grn_global.RData"))
saveRDS(global_node_properties, file.path("results", "bcell_grn_global_node_properties.rds"))
saveRDS(network_summary, file.path("results", "bcell_grn_by_cell_type_summary.rds"))
```
