---
title: "10 - Network Analysis with igraph"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(rtracklayer)
  library(reshape2)
  library(igraph)
  library(ggplot2)
})
```

# Goal

Build GRNs and compute network properties:

- degree: total number of edges connected to a node
- betweenness: influence of a node over information flow in the network
- edges: TF-gene connections weighted by correlation strength
- nodes: TFs and target genes in the network

In this version, Ensembl IDs are converted to gene symbols before network construction.

# Load inputs

```{r load-inputs}
tf_ocr_gene_extended <- readRDS(
  file.path("results", "tf_ocr_gene_interactions_curated_extended.rds")
)

cluster_rna <- readRDS(
  file.path("results", "rna_deg_clustering.rds")
)

rna_anno <- readRDS(
  file.path("data", "rna_gene_annotation.rds")
)

id_map <- rna_anno |>
  dplyr::select(
    ensembl_gene_id,
    hgnc_symbol
  ) |>
  dplyr::distinct()

map_gene_name <- function(x) {
  y <- id_map$hgnc_symbol[
    match(
      x,
      id_map$ensembl_gene_id
    )
  ]

  y[is.na(y) | y == ""] <- x[is.na(y) | y == ""]

  y
}
```

# Helper function

```{r centrality-function}
add_centrality <- function(net) {
  V(net)$degree <- igraph::degree(net)
  V(net)$indegree <- igraph::degree(net, mode = "in")
  V(net)$outdegree <- igraph::degree(net, mode = "out")
  V(net)$eig <- eigen_centrality(net, directed = TRUE)$vector
  V(net)$hubs <- hub_score(net)$vector
  V(net)$authorities <- authority_score(net)$vector
  V(net)$closeness <- closeness(net)
  V(net)$betweenness <- betweenness(net)

  net
}
```

# Global B-cell GRN

```{r global-grn}
tf_gene <- unique(data.frame(
  tf_ensembl = tf_ocr_gene_extended$tf_ensembl,
  gene_ensembl = tf_ocr_gene_extended$gene_ensembl,
  gene_tf_rho = tf_ocr_gene_extended$gene_tf_rho
))

tf_gene$tf_name <- map_gene_name(tf_gene$tf_ensembl)
tf_gene$gene_name <- map_gene_name(tf_gene$gene_ensembl)

edges <- unique(data.frame(
  from = tf_gene$tf_name,
  to = tf_gene$gene_name,
  weight = abs(tf_gene$gene_tf_rho)
))

edges <- edges[
  !is.na(edges$from) &
    !is.na(edges$to) &
    edges$from != "" &
    edges$to != "",
]

nodes <- data.frame(
  id = unique(c(edges$from, edges$to)),
  label = unique(c(edges$from, edges$to))
)

net <- graph_from_data_frame(
  d = edges,
  vertices = nodes,
  directed = TRUE
)

net <- add_centrality(net)

net
```

# Node properties

```{r global-node-properties}
global_node_properties <- data.frame(
  node = V(net)$name,
  degree = V(net)$degree,
  indegree = V(net)$indegree,
  outdegree = V(net)$outdegree,
  eigenvector = V(net)$eig,
  hub_score = V(net)$hubs,
  authority_score = V(net)$authorities,
  closeness = V(net)$closeness,
  betweenness = V(net)$betweenness
)

global_node_properties |>
  arrange(desc(outdegree)) |>
  head(20)
```

# Cell-type-specific GRNs

```{r cell-type-networks}
cell_type <- c(
  "HSC",
  "CLP",
  "proB",
  "preB",
  "ImmatureB",
  "Transitional_B",
  "Naive_CD5neg",
  "Naive_CD5pos"
)

list_grn <- vector("list", length(cell_type))
names(list_grn) <- cell_type

for (ct in cell_type) {

  interaction_col <- paste0(ct, "_interaction")

  aa <- tf_ocr_gene_extended[
    tf_ocr_gene_extended[[interaction_col]] == "yes",
  ]

  if (ct %in% colnames(cluster_rna)) {

    exp_genes <- rownames(
      cluster_rna[
        cluster_rna[[ct]] == 1,
        ,
        drop = FALSE
      ]
    )

    aa <- aa[
      aa$tf_ensembl %in% exp_genes &
        aa$gene_ensembl %in% exp_genes,
    ]
  }

  aa$tf_name <- map_gene_name(aa$tf_ensembl)
  aa$gene_name <- map_gene_name(aa$gene_ensembl)

  edges <- unique(data.frame(
    from = aa$tf_name,
    to = aa$gene_name,
    weight = abs(aa$gene_tf_rho)
  ))

  edges <- edges[
    !is.na(edges$from) &
      !is.na(edges$to) &
      edges$from != "" &
      edges$to != "",
  ]

  if (nrow(edges) == 0) {
    list_grn[[ct]] <- make_empty_graph(directed = TRUE)
    next
  }

  nodes <- data.frame(
    id = unique(c(edges$from, edges$to)),
    label = unique(c(edges$from, edges$to))
  )

  ct_net <- graph_from_data_frame(
    d = edges,
    vertices = nodes,
    directed = TRUE
  )

  ct_net <- add_centrality(ct_net)

  list_grn[[ct]] <- ct_net
}

list_grn
```

# Compare networks

```{r network-summary}
network_summary <- data.frame(
  cell_type = names(list_grn),
  nodes = sapply(list_grn, vcount),
  edges = sapply(list_grn, ecount)
)

network_summary
```

# Top regulators per cell type

```{r top-regulators-per-cell-type}
top_regulators_by_cell_type <- purrr::map_dfr(
  names(list_grn),
  function(ct) {

    g <- list_grn[[ct]]

    if (vcount(g) == 0) {
      return(data.frame())
    }

    data.frame(
      cell_type = ct,
      node = V(g)$name,
      degree = V(g)$degree,
      indegree = V(g)$indegree,
      outdegree = V(g)$outdegree,
      hub_score = V(g)$hubs,
      betweenness = V(g)$betweenness
    ) |>
      arrange(desc(outdegree)) |>
      slice_head(n = 20)
  }
)

top_regulators_by_cell_type
```

# Save networks

```{r save-networks}
saveRDS(
  list_grn,
  file.path("results", "bcell_grn_by_cell_type.rds")
)

save(
  net,
  file = file.path("results", "bcell_grn_global.RData")
)

saveRDS(
  global_node_properties,
  file.path("results", "bcell_grn_global_node_properties.rds")
)

saveRDS(
  network_summary,
  file.path("results", "bcell_grn_by_cell_type_summary.rds")
)

saveRDS(
  top_regulators_by_cell_type,
  file.path("results", "bcell_grn_top_regulators_by_cell_type.rds")
)
```
